---
title: "Bird Multiverse"
format: html
editor: visual
---

## Introduction
In the visualization, I build a “Bird Multiverse” of migratory species by plotting every bird at its geographic centroid on a global map and then highlighting each habitat in turn. A geographic K-nearest-neighbors network connects species by spatial proximity, while node aesthetics encode key life-history traits: point size represents range size, opacity corresponds to body mass, shape differentiates trophic levels, and fill color marks habitat. The most broadly distributed species in each habitat is labeled for easy reference.To ensure even representation and manage computational load, I sampled one species per family (83 families in total). The life-history data are drawn from the AVONET database.

## Load Packages

```{r}
library(tidyverse)
library(FNN)
library(scales)
library(ggrepel)
library(RColorBrewer)
library(rnaturalearth)
library(sf)
```

## Visulization Code
### Data processing
```{r}
# load bird avonet dataset
d <- read.csv("https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/AVONETdataset1.csv")

# trim data
d <- d %>%
  select(-c("Traits.inferred", "Reference.species", "Mass.Refs.Other"))
d <- na.omit(d)

# select migratory birds only
d1 <- d %>%
  filter(Migration == 3)

# pick one species from each family
set.seed(42)  
df <- d1 %>%
  group_by(Family1) %>%
  sample_n(1) %>%
  ungroup()
```

### Visualization preparation
```{r}
# load word map
world <- ne_countries(scale = "medium", returnclass = "sf")

# clean and standardize data for visualization
df_graph <- df %>%
  mutate(Species1 = str_trim(Species1)) %>%
  filter(!is.na(Species1), !is.na(Habitat),
         !is.na(Mass), !is.na(Range.Size),
         !is.na(Centroid.Longitude), !is.na(Centroid.Latitude),
         !is.na(Trophic.Level)) %>%
  mutate(
    name       = Species1,
    Mass_alpha = rescale(Mass,       to = c(0.3, 1)),
    Range_size = rescale(Range.Size, to = c(2, 8))
  )

# build global Geo KNN edge based on centroid lat/lon of habitat range
coords_all <- df_graph %>% select(x = Centroid.Longitude, y = Centroid.Latitude) %>% as.matrix()
knn_all    <- get.knn(coords_all, k = 3) # k = 3, connect to the closet three points
edges_all  <- tibble(
  from = rep(df_graph$name, each = 3),
  to   = as.vector(df_graph$name[knn_all$nn.index])
) %>% 
  filter(from != to) %>% distinct()

edges_all2 <- edges_all %>%
  left_join(df_graph %>% select(name, x = Centroid.Longitude, y = Centroid.Latitude),
            by = c("from" = "name")) %>%
  left_join(df_graph %>% select(name, xend = Centroid.Longitude, yend = Centroid.Latitude),
            by = c("to"   = "name"))

# get coords
edges_all2 <- edges_all %>%
  left_join(df_graph %>% select(name, x = Centroid.Longitude, y = Centroid.Latitude),
            by = c("from" = "name")) %>%
  left_join(df_graph %>% select(name, xend = Centroid.Longitude, yend = Centroid.Latitude),
            by = c("to"   = "name"))

# define color and shape for different habitat and trophic level
habitats   <- unique(df_graph$Habitat)
pal_base   <- brewer.pal(min(length(habitats), 8), "Set2")
if(length(habitats) > length(pal_base)) 
  pal_base <- colorRampPalette(pal_base)(length(habitats))
hab_pal    <- setNames(pal_base, habitats)

trophic_lvls <- unique(df_graph$Trophic.Level)
fill_shapes   <- c(21,22,23,24,25,0,1,2) 
shape_map     <- setNames(fill_shapes[1:length(trophic_lvls)], trophic_lvls)
```

### Visualization
```{r}
p_base_map <- ggplot(data = world) +
  geom_sf(fill = "gray95", color = "gray80") +
  geom_segment(data = edges_all2,
               aes(x = x, y = y, xend = xend, yend = yend),
               color = "grey80", alpha = 0.2) +
  geom_point(data = df_graph,
             aes(x = Centroid.Longitude, y = Centroid.Latitude),
             color = "grey70", size = 1.5, alpha = 0.4) +
  coord_sf(expand = FALSE) +
  theme_void() +
  ggtitle("Bird Multiverse Base Network\n(All Migratory Families)") +
  theme(plot.title = element_text(hjust = 0.5))

# print global map
print(p_base_map)

# highlight overlay of each habitat
plots_map <- map(habitats, function(hab) {
  sub <- df_graph %>% filter(Habitat == hab)
  n   <- nrow(sub)
  
  ggplot(data = world) +
    # global map
    geom_sf(fill = "gray95", color = "gray80") +
    # base network (built from knn)
    geom_segment(data = edges_all2,
                 aes(x = x, y = y, xend = xend, yend = yend),
                 color = "grey80", alpha = 0.1) +
    geom_point(data = df_graph,
               aes(x = Centroid.Longitude, y = Centroid.Latitude),
               color = "grey70", size = 1, alpha = 0.3) +
    # highlight current habitat node
    geom_point(data = sub,
               aes(x = Centroid.Longitude, y = Centroid.Latitude,
                   size  = Range_size,
                   alpha = Mass_alpha,
                   shape = Trophic.Level),
               fill = hab_pal[hab],
               color = "black",
               stroke = 0.6) +
    # label the species with the maximum distribution range
    geom_text_repel(data = sub %>% slice_max(Range_size, n = 1),
                    aes(x = Centroid.Longitude, y = Centroid.Latitude, label = name),
                    size = 2) +
    coord_sf(expand = FALSE) +
    ggtitle(glue::glue("{hab} (n = {n})")) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5),
          legend.position = "bottom") +
    scale_size_continuous(range = c(2,8),  guide = guide_legend("Range Size")) +
    scale_alpha_continuous(range = c(0.3,1),guide = guide_legend("Mass")) +
    scale_shape_manual(values = shape_map,   guide = guide_legend("Trophic Level")) +
    guides(
    size  = guide_legend("Range Size",   nrow = 2, byrow = TRUE),
    alpha = guide_legend("Mass",         nrow = 2, byrow = TRUE),
    shape = guide_legend("Trophic Level",nrow = 2, byrow = TRUE)
  ) +
  theme(
    legend.position = "bottom",
    legend.box      = "horizontal"
  )
})

names(plots_map) <- habitats

# print map for each habitat
for(h in habitats) print(plots_map[[h]])
```